A GPU Sonar Simulator for Automatic Target Recognition

### 2014-05-28 Simulator/article.tex
More simulator

\ifPeerReview
\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
\else
\documentclass[journal]{IEEEtran}
\fi

@@

\RequirePackage{array}
% 
% \ifCLASSOPTIONcompsoc
%    \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
% \else
%    \usepackage[caption=false,font=footnotesize]{subfig}
% \fi

@@

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}. However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b} which, similar to our approach, takes advantage of parallel processing at GPUs and uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.

@@

The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane. This is achieved by applying a set of transformations to each vertex in the model. In OpenGL, the set of transformations are usually formalized as
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
\end{align*}
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices for scaling, rotation and translation, respectively. These are described in Appendix \ref{transformation_matrices}.

Next we map the model into world coordinates using the view matrix $\V$. This involves positioning and orienting the model relative to the camera. In our case we let the camera be the phase center of the sonar, and define it as the world origin. The transformation take the form of
\begin{align}
\V &= \mat{T}(0, dy, dz) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  0 \\
0  &  1  &  0  &  dy \\
0  &  0  &  1  &  dz \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
where
\begin{align*}
dy &= \\
dz &= \\
\phi_x &= \arctan\left(\frac{y_\text{mean}}{x_\text{mean}}\right)
\end{align*}

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. Its projection matrix is defined as
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene. We need these parameters set They define the window through which we observe our scene. We want this adapted to perfectly fit the specified image size. This is straightforward for left and right since it matches the $x$-axis
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
For the near and far limits one simply needs to make sure the scene will fit in it
\begin{align}
n &= 0\;\text{m}   &  f &= 300\;\text{m} 
\end{align}
How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
and the distance from the image origin to its $y$-boundaries is given as
\begin{align}
\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
\end{align}
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
\begin{align}
\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
\end{align}

@@

\begin{figure}[p]\centering%
\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
\caption{SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.}\label{data_cylinder}%
\end{subfigure}
\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
\caption{Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.}\label{sim_cylinder}%
\end{subfigure}
\caption{A SAS image of a cylinder and a template simulation adapted to it.}
\end{figure}

\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}
\begin{figure}[p]\centering%
\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
\caption{Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.}\label{overlay_cylinder}%
\end{subfigure}\hspace{\imgspacing}%
\begin{subfigure}[t]{.5\linewidth-\imgspacing/2}
\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
\caption{Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.}\label{overlay_cylinder_adaptive}%
\end{subfigure}
\hfill\parbox{\linewidth-0pt}{\small\centering
Segmentation overlay color description:\\[.5\baselineskip]
\begin{minipage}{.49\linewidth}
\cdesc{800000}{Template highlight and image highlight}
\cdesc{FF1000}{Template highlight only}
\cdesc{FFEB00}{Image highlight only}
\cdesc{83FF7C}{Background pixels}
\end{minipage}%
\begin{minipage}{.49\linewidth}
\cdesc{000083}{Template shadow and image shadow}
\cdesc{0014FF}{Template shadow only}
\cdesc{00EFFF}{Image shadow only}
\cdesc{00A7FF}{Template shadow and image highlight}
\end{minipage}}
\vspace{\baselineskip}
\caption{Segmented simulation image overlaid the segmented SAS image.}
\end{figure}

@@

>>>>

\ifPeerReview
\documentclass[10pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
\else
\documentclass[journal]{IEEEtran}
\fi

@@

\RequirePackage{array}

\ifCLASSOPTIONcompsoc
   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
   \usepackage[caption=false,font=footnotesize]{subfig}
\fi

@@

Various simulators for high frequency, side-looking sonar imagery have been published, e.g.~\cite{Bell1997,Sammelm2003}. However, most implementations have prioritized accurate acoustic modeling at the expense of execution speed. One exception is the SIGMAS+ simulator in~\cite{Coiras2009a, Coiras2009b}. Similarly to our approach it takes advantage of parallel processing on GPUs, and it uses OpenGL to render a 2D image from a 3D model, under the assumption of Lambertian scattering. However, SIGMAS+ is still aimed more towards realistically looking images, including effects like ambient noise, which is irrelevant for our purpose of template generation. Also, their simulator sums images obtained from multiple OpenGL rendering passes to create the sonar image, while we render once and post-process this result with OpenCL to tailor the sonar image with more flexibility.

@@

The next step is to scale, orient and position the objects in the scene, and finally project this scene onto an image plane. This is achieved by applying a set of transformations to each vertex in the model. In OpenGL, the set of transformations can be formalized as
\begin{align*}
\bmat{x\\y\\0\\1}_\text{image} = \P \cdot \V \cdot \M \cdot \bmat{x\\y\\z\\1}_\text{model},
\end{align*}
where $\bmat{\cdot}_\text{model}$ is the loaded model, $\bmat{\cdot}_\text{image}$ is the output image, and $\M$, $\V$ and $\P$ are transformation matrices. These will be described next.

The first transform we apply is the model matrix $\M$, which operates in the model's local coordinate system. It is defined as
\begin{align}
\bmat{x_t \\ y_t \\ z_t \\ w}_\text{\parbox{1cm}{\setlength\baselineskip{0.3cm}transformed\\model}} &= \boldsymbol{T} \cdot \R \cdot \boldsymbol{S} \bmat{x \\ y \\ z \\ w}_\text{model},
\end{align}
where $\boldsymbol{S}, \R$ and $\boldsymbol{T}$ are matrices\todo{be explicit on dependencies?} for scaling, rotation and translation, respectively. These are provided for reference in Appendix \ref{transformation_matrices}.

Next we map the model into world coordinates using the view matrix $\V$. This involves aligning and positioning the model into the view of the camera. In our case the camera represents the phase center of the sonar, and it is placed in the world origin. The transformation take the form of
\begin{align}
\V &= \mat{T}(x_\text{c}, y_\text{c}, z_\text{c}) \cdot \R_\text{x}(\phi_x, 0, 0) \nonumber\\
&= 
\bmat{
1  &  0  &  0  &  x_\text{c} \\
0  &  1  &  0  &  y_\text{c} \\
0  &  0  &  1  &  z_\text{c} \\
0  &  0  &  0  &  1 \\
} \cdot
\bmat{
1  &  0           &  0           &  0 \\
0  &  \cos\phi_x  &  -\sin\phi_x &  0 \\
0  &  \sin\phi_x  &  \cos\phi_x  &  0 \\
0  &  0           &  0           &  1 \\
},
\end{align}
where $(x_\text{c}, y_\text{c}, z_\text{c})$ is the image center coordinates for along track, cross track, and depth, respectively, and $\Phi_x$ is the elevation angle\todo{is this correct?} of the sonar. It is given as
\begin{align*}
\phi_x &= \arctan\left(\frac{y_\text{mean}}{-z_\text{mean}}\right).
\end{align*}\todo{use absolute sign?}

Finally the relevant part of the scene is projected onto the image plane with a projection matrix $\P$. There are two main types of projection, perspective and orthographic. Perspective projection scales all $x$ and $y$ coordinates inversely with the distance to the objects to create a sense of perspective in the image. Orthographic projection, on the other hand, maps all the scene's vertices to a squared cube where each dimension has the range [-1,1]. This will render all objects equally large whether they are in the distance or up close. The matrix that gives orthographic projection is defined as
\begin{align}
\P = \left(\begin{matrix}
\frac{2}{r-l}  &  0              &  0              &  -\frac{r+l}{r-l} \\
0              &  \frac{2}{t-b}  &  0              &  -\frac{t+b}{t-b} \\
0              &  0              &  \frac{-2}{f-n} &  -\frac{f+n}{f-n} \\
0              &  0              &  0              &  1
\end{matrix}\right)
\end{align}
where $l,r,b$ and $t$ are the left, right, bottom and top boundaries of the projection plane, respectively, and where $n$ and $f$ is the near and far limits of the cube that encompass the scene. These parameters need to be adapted to provide a 2D image that matches the image size, which we specify in along track and cross track coordinates ($x,y$). This is straightforward for left and right since matches the $x$-axis
% They define the window through which we observe our scene.
\begin{align}
l &= x_\text{min}  &  r &= x_\text{max}
\end{align}
For the near and far limits one simply needs to make sure the scene will fit in it
\begin{align}
n &= 0\;\text{m}   &  f &= 300\;\text{m} 
\end{align}
How to set the bottom and top bounaries can be inferred from \Fig{camera_top_bottom}. First we define an origin vector that points to the center of the image
\begin{align}
\vec o &= \left[\frac{x_\text{min}+x_\text{max}}{2}, \frac{y_\text{min}+y_\text{max}}{2}, \frac{z_\text{min}+z_\text{max}}{2} \right]\T,
\end{align}
and the distance from the image origin to its $y$-boundaries is given as
\begin{align}
\Delta y &= \frac{y_\text{max} - y_\text{min}}{2}.
\end{align}
To find the nearest and farthest point in the image that will be part of the projected scene, $\alpha$ and $\beta$, we must consider the seafloor rotation angle around the $x$-axis, $\Phi_x$ 
\begin{align}
\alpha &= \vec 0 - \left[0, \Delta y, \Delta y \times \arctan\Phi_\text{x} \right]
\end{align}

@@

\begin{figure}[p]\centering%
\subfloat[SAS image of a cylinder with 2.6\;m length and 0.53\;m radius.]{%
\graphicsAI[drawing,width=\linewidth]{gfx/data_cylinder_submerged.svg}%
\label{data_cylinder}}%
\subfloat[Cylinder simulation with parameters estimated from SAS image: Length 2.6\;m, burial depth 0.263\;m, aspect angle 105$^\circ$.]{
\graphicsAI[drawing,width=\linewidth]{gfx/sim_cylinder_submerged_adaptive.svg}%
\label{sim_cylinder}}%
\caption{A SAS image of a cylinder and a template simulation adapted to it.}
\end{figure}

\newcommand\cdesc[2]{{\raggedright\setlength\fboxsep{0pt}
\fbox{\colorbox[HTML]{#1}{\vrule height8.5pt depth3.5pt width0pt\hspace{.5cm}}}\ \ #2\\}}

\begin{figure}[p]\centering%
\subfloat[Best template out of a predefined set. Inaccurate cylinder length, alignment and immersion depth. The segmented shadow and highlight regions are misaligned.]{
\graphicsAI[drawing,width=\linewidth]{gfx/overlay_cylinder_submerged.svg}%
\label{overlay_cylinder}}%
% \hfil%
\subfloat[Adapted template. Template simulated with adapted parameters. Good fit in both highlight and shadow segments.]{%
\graphicsAI[width=\linewidth]{gfx/overlay_cylinder_submerged_adaptive.svg}%
\label{overlay_cylinder_adaptive}}%
\hfill\parbox{\linewidth}{\small\centering
Segmentation overlay color description:\\[.5\baselineskip]
\begin{minipage}{\linewidth}
\cdesc{800000}{Template highlight and image highlight}
\cdesc{FF1000}{Template highlight only}
\cdesc{FFEB00}{Image highlight only}
\cdesc{83FF7C}{Background pixels}
\end{minipage}%
\begin{minipage}{\linewidth}
\cdesc{000083}{Template shadow and image shadow}
\cdesc{0014FF}{Template shadow only}
\cdesc{00EFFF}{Image shadow only}
\cdesc{00A7FF}{Template shadow and image highlight}
\end{minipage}}
\vspace{\baselineskip}
\caption{Segmented simulation image overlaid the segmented SAS image.}
\end{figure}

@@

\begin{itemize}
\item GPUs low power. Good for integration in AUVs.
\item Good accuracy even with a real-time constraint in e.g. a AUV.
\item Improved inertial navigation? 
\end{itemize}
